Distortion-corrected kernel density estimate on Riemannian manifolds

### 2022-07-19 paper/paper.Rmd
Add reference for intRinsic

<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$. The underlying manifold $M$ has the true density $f$. Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$. The Riemannian metric of the embedding is denoted as $g_i, i = 1, 2, \dots, N$. In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$. -->

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data. @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data. 
In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathcal{R}^d$
<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density $\hat{f}$. KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.

@@

As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned. 
To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process. @Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl]. By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated. An *R* library named *intRinsic* has implemented this algorithm and we are going to apply it to when $d$ is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters $MANILEARN$}
  \BlankLine
  \begin{algorithmic}[1]

@@

  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm $MANILEARN$;

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator). The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \@ref(alg:kdeanomaly). 
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation. Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator). The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f_N(\pmb{y}_i)$.

>>>>

<!-- Denote the $s$-dimensional input data with observations $X_i, i = 1, 2, \dots, N$, where $x_i \in \mathcal{R}^s$. The underlying manifold $M$ has the true density $f$. Manifold learning takes the input data and finds its $d$-dimensional representation of data that lie on the manifold $M$, with output data points $Y_i, i = 1, 2, \dots, N$, where $Y_i \in \mathcal{R}^d$ and $d \ll s$. The Riemannian metric of the embedding is denoted as $H_i, i = 1, 2, \dots, N$. In kernel density estimate, the pointwise variable bandwidth matrix $H_i$ is used to estimate the density of the embedding, $\hat{f}$. -->

@@

Multivariate kernel density estimation has gained lots of attention in exploratory data analysis. It is a non-parametric technique to estimate the data density based on weighted kernels centered at the data. @Scott2015-vl covers a wide range of techniques for estimating the kernel densities of multivariate data. 
In general, given a multivariate random sample $\pmb{X}_1, \dots, \pmb{X}_N \in \mathbb{R}^d$
<!-- $\pmb{X}_i=(X_{i1}, X_{i2}, \dots, X_{id})^\prime \in \mathcal{R}^d$, $i = 1, 2, \dots, N$  -->
drawn from a density $f$, the variable kernel density estimate [@Breiman1977-qc; @Jones1990-oe; @Terrell1992-ut] of $f$ at point $\pmb{x}=(x_1, x_2, \dots, x_d)^\prime \in \mathcal{R}^d$ is given by
\begin{equation}
\label{eq:vkde}
\hat{f}(\pmb{x};\pmb{H}_i)=\frac{1}{N}\sum\limits_{i=1}^N K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i).
\end{equation}
where $K(\pmb{x})$ is the kernel function, $\pmb{H}_i$ is called the bandwidth matrix and it is a symmetric positive definite matrix of dimension $d\times d$, and $K_{\pmb{H}_i}(\pmb{x}) = |\pmb{H}_i|^{-1/2} K(\pmb{H}_i^{-1/2}\pmb{x})$.
When using the standard gaussian kernel, we have $K(\pmb{x}) = (2\pi)^{-d/2}\exp(-\frac{1}{2}\pmb{x}'\pmb{x})$ and hence
\begin{equation}
\label{eq:kernel}
K_{\pmb{H}_i}(\pmb{x}-\pmb{X}_i)=(2\pi)^{-d/2}|\pmb{H}_i|^{-1/2}\exp\left[-\frac{1}{2}(\pmb{x}-\pmb{X}_i)'\pmb{H}_i^{-1}(\pmb{x}-\pmb{X}_i)\right],
\end{equation} 
with $|\pmb{H}_i|$ being the determinant of matrix $\pmb{H}_i$.
Intuitively, KDE smooths out each data point $\pmb{X}_i$ into a smooth bump with the shape determined by the kernel function. Then an average over all data points is taken as the estimated density $\hat{f}$. KDE is a local smooth estimator because the nearby points contributes more to the density value compared to points far away from the estimated data point $\pmb{x}$.

@@

As pointed out by @Perrault-Joncas2013-pq, if the embedding dimension is larger than the manifold intrinsic dimension $d$, the rank of the embedding metric $\pmb{H}(p)$ is equal to $d$; otherwise, the Riemannian metric will be returned. 
To ensure that the estimated Riemannian metric is returned as a $d\times d$ matrix, for unknown manifold dimension, we apply the intrinsic dimension (id) selection algorithm called *two-nearest neighbor estimator (TWO-NN estimator)* [@Facco2017-rl; @Denti2021-jl] to derive id estimator from a Poisson point process. @Facco2017-rl point out that of the distances between a point and its second and first neareast neighbors follow a Pareto distribution with scale parameter $1$ and shape parameter $d$, the intrinsic dimension of the manifold [@Facco2017-rl]. By fitting a Pareto distribution to proper transformed data, the intrinsic dimension could be estimated. An *R* library named *intRinsic* [@Denti2021-qc] has implemented this algorithm and we are going to apply it to when $d$ is unknown.

\begin{algorithm}[!htb]
  \caption{Learn metric algorithm in \cite{Perrault-Joncas2013-pq} }
  \label{alg:learnmetric}
  \DontPrintSemicolon
  \SetAlgoLined
  \SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}\SetKwInOut{Parameter}{parameter}\SetKwInOut{OptParameter}{optimization parameter}
  \Input{ high-dimensional data $\pmb{x}_i \in \mathbb{R}^s$ for all $i=1,\ldots,N$ }
  \Output{ low-dimensional data $\pmb{y}_i \in \mathbb{R}^d$ and its Riemannian metric $\pmb{H}(\pmb{y}_i)$ for all $i=1,\ldots,N$ }
  \Parameter{ embedding dimension $d$, bandwidth parameter $\sqrt{\varepsilon}$, manifold learning algorithm }
  \OptParameter{ manifold learning parameters EMBED }
  \BlankLine
  \begin{algorithmic}[1]

@@

  \STATE Embed all data point $\pmb{X}\in \mathbb{R}^s$ to embedding coordinates $\pmb{Y}=(\pmb{y}^1,\dots,\pmb{y}^d)^\prime$ by any existing manifold learning algorithm EMBED;

@@

Now we are able to generate kernel density estimates of the underlying manifold using the estimator in Equation \@ref(eq:denestimator). The densities could be further applied to unsupervised learning tasks such as clustering and anomaly detection. As one way to utilize the kernel density estimates, we propose an algorithm for anomaly detection in \@ref(alg:kdeanomaly). 
There are mainly four steps involved in the algorithm. The first step is to apply the Learn metric algorithm described in \@ref(alg:learnmetric) to the input high-dimensional data $x_i,i=1,\ldots,N$ where $x_i\in \mathcal{R^s}$ to get the low-dimensional embedding $y_i$, where $y_i\in \mathcal{R^d}$ and the Riemannian metric $\pmb{H}_i$ with dimension $d\times d$ for each observation. Two parameters $\sqrt{\varepsilon} = 0.4$ and $c=0.25$ are given for Gaussian kernels. Then we could use the pointwise Riemannian metric to calculate the kernel density estimate with our proposed estimator in Equation \@ref(eq:denestimator). The top outliers of size $n\_outliers$ are obtained by ordering the embedding points $y_i$ according to their density estimates $f(\pmb{y}_i)$.
